---
title: "Zwin Nature Park (ZNP) Bird Census Counts: Data Processing for DarwinCore mapping"
author: "Stijn Cooleman"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
# Set global options: eval = FALSE could be set by default so the document compiles 
# even if the raw Excel data files are not present in the current working directory.
knitr::opts_chunk$set(
  echo = TRUE, 
  message = TRUE, 
  warning = FALSE
)
```

## Introduction

This R Markdown document contains the data processing pipeline for the [Zwin Nature Park (ZNP) Bird census counts dataset on GBIF](https://www.gbif.org/dataset/dde71542-ad2d-4ec7-a93c-eb18bc0f432b). The pipeline processes raw bird counts and maps essential data according to DarwinCore standards for the occurrences and events, and structures the monitoring metadata in the Humboldt Extension (HE).

### Load Libraries

We load the `tidyverse` suite (which includes `readr`, `dplyr`, `tidyr`, etc.) and `readxl` for importing Excel files.

```{r load-libraries}
library(tidyverse)    # To do data science (includes readr)
library(readxl)       # To read excel sheets
```

---

## Part 1: Occurrence Extension

This section parses and cleans the bird count observation spreadsheets for 2020: Inner Trail, Transects, and Point counts.

### 1.1 Data from Inner Trail Counts

First, we read the Inner Trail (IT) counts Excel sheet. (Note: empty cells were manually replaced with zeros in Excel for this sheet).

```{r read-inner-trail}
## Read data file
zp20IT <- readxl::read_xlsx("./2025-02-20/InnerTrail20_over+0_155sp_R.xlsx")
```

Next, we reshape the data from wide to long format using `pivot_longer` (excluding the December-B event since there was no counting in the last half of December 2020).

```{r pivot-inner-trail}
occ_zp20IT <- zp20IT %>% 
  pivot_longer(
    cols = c('ZNP:Birds:2020-01-A:IT':'ZNP:Birds:2020-12-A:IT'),
    names_to = "eventID",
    values_to = "nr_obs",
    values_drop_na = TRUE
  )
```

We now add standard occurrence attributes based on the observations: 
* occurrenceStatus: Set as `'present'` if `nr_obs > 0`, otherwise `'absent'`. 
* occurrenceRemarks: Set as `'flying over'` if `nr_obs == 'over'`. 
* individualCount: Set as numeric value, coercing `'over'` and `'x'` to `NA`. 

```{r process-inner-trail}
# Add occurrenceStatus
occ_zp20IT_occstatu <- occ_zp20IT %>% 
  dplyr::mutate(occurrenceStatus = ifelse(nr_obs > 0, 'present', 'absent'))

# Add occurrenceRemarks
occ_zp20IT_sr <- occ_zp20IT_occstatu %>% 
  dplyr::mutate(occurrenceRemarks = ifelse(nr_obs == 'over', 'flying over', NA))

# Add individualCount and remove temporary nr_obs column
occ_zp20IT_sri <- occ_zp20IT_sr %>% 
  dplyr::mutate(individualCount = ifelse(nr_obs %in% c('over','x'), NA, nr_obs))

# Select all except nr_obs
occ_zp20IT_b <- occ_zp20IT_sri %>% 
  select(-nr_obs)

# Convert individualCount to numeric
occ_zp20IT_fin <- occ_zp20IT_b %>% 
  mutate(individualCount = as.numeric(individualCount))
```

---

### 1.2 Data from Transect Counts

Next, we process the 2020 short transect counts data. Here, we programmatically replace `NA` values with `0`.

```{r read-and-process-transects}
# Read short transects data file
zp20T <- readxl::read_xlsx("./2025-02-20/Transects20_109sp_R.xlsx")

# Replace NA with 0 programmatically
occ0_zp20T <- zp20T %>% 
  mutate_if(~ any(is.na(.x)), ~ if_else(is.na(.x), 0, .x))

# Reshape transects data to long format
occ_zp20T <- occ0_zp20T %>% 
  pivot_longer(
    cols = c('ZNP:Birds:2020-01-A:FT':'ZNP:Birds:2020-12-A:DT'),
    names_to = "eventID",
    values_to = "individualCount",
    values_drop_na = TRUE
  )

# Add occurrenceStatus based on individualCount
occ_zp20T_fin <- occ_zp20T %>% 
  dplyr::mutate(occurrenceStatus = ifelse(individualCount > 0, 'present', 'absent'))
```

---

### 1.3 Data from Point Counts

Next, we process the 2020 point counts data.

```{r read-and-process-points}
# Read point counts data file
zp20P <- readxl::read_xlsx("./2025-02-20/Points20_108sp_R.xlsx")

# Replace NA with 0 programmatically
occ0_zp20P <- zp20P %>% 
  mutate_if(~ any(is.na(.x)), ~ if_else(is.na(.x), 0, .x))

# Reshape point counts data to long format
occ_zp20P <- occ0_zp20P %>% 
  pivot_longer(
    cols = c('ZNP:Birds:2020-01-A:FTP':'ZNP:Birds:2020-12-A:DTP'),
    names_to = "eventID",
    values_to = "individualCount",
    values_drop_na = TRUE
  )

# Add occurrenceStatus based on individualCount
occ_zp20P_fin <- occ_zp20P %>% 
  dplyr::mutate(occurrenceStatus = ifelse(individualCount > 0, 'present', 'absent'))
```

---

### 1.4 Combining Count data frames

We combine the data from all three data frames (Inner Trail, Transects, and Point counts) into one data frame.

```{r combine-2020-occurrences}
# Bind the 3 occurrence sheets together, adding an 'id' column to track origin
occ_zp20 <- bind_rows(
  occ_zp20IT_fin, 
  occ_zp20T_fin, 
  occ_zp20P_fin, 
  .id = "id"
)
```

We now add occurrence IDs, detect taxon rank (subspecies, form, or species), and specify default metadata attributes (kingdom, basis of record, and identifier).

```{r finalize-2020-occurrences}
occ_zp2020 <- occ_zp20 %>% 
  # Generate occurrenceID using row numbers
  mutate(internID = row_number()) %>% 
  mutate(occurrenceID = paste("ZNP:Birds:20:", sprintf("%05d", internID), sep = "")) %>% 
  select(-internID) %>% 
  # Determine taxonRank based on naming conventions
  mutate(taxonRank = ifelse(str_detect(vernacularName, pattern = "[(]"), "subspecies",
                            ifelse(str_like(scientificName, "% f. %"),
                                   "form",
                                   "species"))) %>% 
  # Default metadata values
  mutate(kingdom = "Animalia") %>% 
  mutate(basisOfRecord = "HumanObservation") %>% 
  mutate(identifiedBy = "Wouter Faveyts") %>% 
  mutate(identificationVerificationStatus = "validated") %>% 
  select(-id)
```

---

### 1.5 Linking WoRMS LSIDs

To facilitate our data integration within the Ocean Biodiversity Information System (OBIS) network, we map species names to World Register of Marine Species (WoRMS) Life Science Identifiers (LSIDs).

```{r add-worms-lsids}
# Load WoRMS mapping spreadsheet
zp_worms <- read_xlsx("./znp_vogels_csv_sources_ds_v1.6/Worms_lsid_60sp_znp.xlsx")

# Select only necessary columns from WoRMS mappings
zp_worms_1 <- zp_worms %>% 
  select(scientificName, LSID)

# Join the WoRMS LSIDs onto our occurrences dataset
zp_df1 <- occ_zp2020 %>% 
  left_join(zp_worms_1, by = "scientificName")

# Rename LSID column and set verification status
zp_occ <- zp_df1 %>% 
  rename(scientificNameID = LSID)
```

### 1.6 Compiling occurrence file 

```{r occ csv}
# Write finalized occurrence CSV
write_csv(zp_occ, "zp_20_occ.csv",na = "")

# Display the first 50 rows of the finalized occurrence data
head(zp_occ, 50)
```

---

## Part 2: Events Core

This section processes and standardizes counts events, dates, and locations.

### 2.1 Standardizing Event Time

We load raw 2020 event metadata, split/parse start and end times, and construct DarwinCore compliant ISO 8601 interval strings.

```{r process-2020-event-times}
# Read raw 2020 event details
zp20events_t <- readxl::read_xlsx("./2025-02-20/events_all_R_2020_+Time.xlsx")

# 1. Parse Start Hour
# Separate hours (hh) and minutes (mm) from "Startuur" (e.g. "09u15" or "09U15")
zp20events_t_s <- zp20events_t %>% 
  separate_wider_delim(Startuur, regex("u", ignore_case = TRUE), names_sep = "")

# Rename columns
zp20events_t_s_hm <- zp20events_t_s %>% 
  rename(s_hh = Startuur1, s_mm = Startuur2)

# If minutes are blank, fill with "00"
zp20events_t_s_hm$s_mm[zp20events_t_s_hm$s_mm == ""] <- "00"

# 2. Parse End Hour
zp20events_t_se <- zp20events_t_s_hm %>% 
  separate_wider_delim(Einduur, regex("u", ignore_case = TRUE), names_sep = "") %>% 
  rename(e_hh = Einduur1, e_mm = Einduur2)

# If minutes are blank, fill with "00"
zp20events_t_se$e_mm[zp20events_t_se$e_mm == ""] <- "00"

# 3. Time Adjustment
# Convert parts to numeric 
zp20events_t_se1 <- zp20events_t_se %>% 
  mutate(s_mm = as.numeric(s_mm)) %>% 
  mutate(e_mm = as.numeric(e_mm)) %>%   
  mutate(s_hh = as.numeric(s_hh)) %>% 
  mutate(e_hh = as.numeric(e_hh))

# Format hours and minutes back to hh:mm
zp20events_t_se2 <- zp20events_t_se1 %>% 
  mutate(starttime = sprintf("%02d:%02d", s_hh, s_mm)) %>% 
  mutate(endtime = sprintf("%02d:%02d", e_hh, e_mm))

# Merge starttime and endtime into one data field
zp20events_t_se3 <- zp20events_t_se2 %>% mutate(eventTime = paste(starttime,endtime, sep = ""))

# 4. Refine columns
zp20events_t_se4 <- zp20events_t_se3 %>% 
  select(eventDate, eventID, parentEventID, locationID, eventTime, sampleSizeValue, sampleSizeUnit, HOURS)

# Format date column
zp20events_date <- zp20events_t_se4 %>% 
  mutate(eventDate = as.Date(eventDate, format = "%Y-%m-%d"))

# Export intermediate CSVs (retaining or dropping HOURS)
write_csv(zp20events_date, "zp20events_dates&hours.csv")

zp20events_dates <- zp20events_date %>% 
  select(-HOURS)

# Intermezzo notes

## Filter out possible empty parentEventIDs (keep only sub-events in the dwc_term eventID, which was just appropriate for the 2024 spreadsheet)
## zp24_events_date_f <- zp24_events_date %>% 
##   filter(parentEventID != "")
## View(zp24_events_date_f)

## Temporal scope (duration in minutes) for years where durations are stored separately, basically if 'tijdsduur' (esp. in minutes) is not included in "events_all_R_year.xlsx" = was appropriate for spreadsheets from 2022,2023,2024
### Load duration values
### zp_22_t_raw <- readxl::read_xlsx("./2025-02-20/events_22_tijdsduur.xlsx")
### # summary(zp_22_t_raw)
```

---

### 2.2 Spatial Details Joining

We merge spatial details (coordinates, locality name, coordinate uncertainty in meters) into the events dataset.

```{r join-localities}
# Load location data, incl. locality reference coordinates (with refined uncertainty radius values)
zp_loc <- readxl::read_xlsx("./2025-02-20/Localities_events_adj.xlsx")

# Join adjusted localities onto 2020 events
zp_20_events <- zp20events_dates %>% 
  left_join(zp_loc, by = "locationID")
```

### 2.3 Compiling events file

```{r eve csv}

# Write finalized events core CSV
write_csv(zp_20_events, "zp_20_events.csv",na = "")

# Display the first 50 rows of the finalized events data
head(zp_20_events, 50)
```

---

## Part 3: Humboldt Extension (HE)

The Humboldt Extension captures standardized monitoring metadata, such as sampling protocol efforts (such as person-hours), and relates specific count events to their survey metadata.

```{r process-humboldt-extension}

# 1. Select columns of interest for the Humboldt Extension from the events metadata
zp_20_events_sel <- zp_20_events %>% 
  select(eventID, locationID, sampleSizeValue, sampleSizeUnit, samplingProtocol)

# 2. Read parent event IDs reference file
zp_20_paev_HE_b <- readxl::read_xlsx("./2025-02-20/zp_20_parentEventIDs.xlsx")

# Add a fictive locationID indicator ('paev') to mark them as parent events
zp_20_paev_HE_b$locationID <- 'paev'

# 3. Combine event levels and parent-event levels
zp_20_HE_b <- bind_rows(zp_20_events_sel, zp_20_paev_HE_b)

# 4. Read standard details/exceptions details file for HE
zp_HE_d <- readxl::read_xlsx("./2025-02-20/zp_HE_d.xlsx")

# 5. Join Humboldt details on locationID
zp_20_HE_c <- zp_20_HE_b %>% 
  left_join(zp_HE_d, by = "locationID")

# 6. Calculate person-hours from minutes (sampleSizeValue) and round to 2 significant digits
zp_20_HE_f <- zp_20_HE_c %>% 
  mutate(samplingEffortValue = signif(sampleSizeValue / 60, digits = 2))

# Note: Alternative formats could be constructed using sprintf:
# e.g., mutate(samplingEffortValue = sprintf("%01d:%02d", sampleSizeValue / 60))

# 7. Drop unnecessary temporary columns
zp_20_HE <- zp_20_HE_f %>% 
  select(-c(locationID, sampleSizeValue, sampleSizeUnit, samplingProtocol))

# Write finalized Humboldt Extension CSV
write_csv(zp_20_HE, "zp_20_HE.csv",na = "")

# Display the first 50 rows of the finalized HE extension data
head(zp_20_HE, 50)
```
